---
title: "To export or not export `phruta` outputs"
author: Cristian Román-Palacios
date: '`r paste("First created on 2021-07-30. Updated on", Sys.Date())`'
vignette: >
  %\VignetteIndexEntry{To export or not export `phruta` outputs}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
output:
#  pdf_document:
#    highlight: null
#    number_sections: yes
  knitr:::html_vignette:
    toc: true
    fig_caption: yes
---

```{r include = FALSE}
knitr::opts_chunk$set(eval = TRUE, warning = FALSE, message = FALSE, progress = FALSE)
unlink(c("0.Sequences", "1.CuratedSequences", "2.Alignments"), recursive = TRUE)
start_time <- Sys.time()
library(knitr)
```

# Table of Contents

1.  [To export or not export from `phruta`](#intro)
2.  [From taxonomic names to sequence alignments, exporting](#paragraph1)

## To export or not export from `phruta` <a name="intro"></a>

Assembling a molecular dataset for particular target taxa with `phruta` can be performed almost entirely by saving objects into your workspace. This topic was covered in the introductory vignette to `phruta` (Using the `phruta` R package). However, in some situations, it might be desirable to save the outputs of different `phruta` functions to particular folders. This tutorial will cover that specific situation. Specifically, we will be reviewing how `phruta` can be used to export `.fasta` and `.csv` files that are generated in different steps of the pipeline.

## From taxonomic names to sequence alignments, exporting data <a name="paragraph1"></a>

In the introductory vignette to `phruta` (Using the `phruta` R package), we assembled a basic molecular dataset for inferring the phylogeny among three mammal genera. Let's recreate the same tutorial, but in this case, exporting the intermediate files that are created after using several of the functions. Note that the structure of this tutorial closely follows that of "Using the `phruta` R package". You should be able to follow this tutorial even without having reviewed the introductory vignette.

Please assume that we are interested in building a phylogenetic tree for the following three genera: *Felis, Vulpes*, and *Phoca*. All these three genera are classified within the Carnivora, a mammalian order. Both *Felis* and *Vulpes* are classified in different superfamilies within the Fissipedia. Finally, *Phoca* is part of another suborder, Pinnipedia. We're going to root our tree with another mammal species, a Chinese Pangolin (*Manis pentadactyla*). Users can select additional target species and clades. However, for simplicity, we will run the analyses using three genera in the ingroup and a single outgroup species.

So far, we have decided the taxonomic make of our analyses in `phruta`. We will also need to determine the gene regions to be used in our analyses. Fortunately, mammals are extensively studied and a comprehensive list of potential gene regions to be analyzed is already available. For instance, we could use same gene regions sampled in [Upham et al (2009)](https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.3000494#sec030). However, for this tutorial, we will simply try to find the gene regions are well sampled for the target taxa. I believe that figuring out the best sampled gene regions in genbank, instead of providing gene names, is potentially more valuable when working with poorly studied groups (e.g. invertebrates). Before we move on, please make sure that you you have [set a working directory for this project](http://rfunction.com/archives/1001). All the files will be saved to this directory. `phruta` will politely ask before writing files to your local directories.

Let's start by loading `phruta`!

```{r eval = TRUE, warning = TRUE, message = TRUE, progress = TRUE}
library(phruta)
```

```{r setup, echo=FALSE, include = FALSE}
##This indicates that the user doesn't need to confirm (y) before phruta writes to their local machine. This is only for testing purposes...
assign(".testMode", TRUE, envir = phruta:::pkg.env)
```

Now, let's look for the gene regions that are sampled for our target taxa. Again, this step is not always necessary. In some groups, gene-level sampling is very standard (e.g. COI, 12S). However, the structure of gene sampling sometimes becomes more blurry as you zoom out taxonomically. For instance, genes A and B can be extensively sampled in genus 1. However, genus 2 in the same family has mainly been studied using genes Y and Z. The idea here is that `phruta` will try to find those gene regions that are extensively sampled across species in the target taxa. We will use the `gene.sampling.retrieve` function in `phruta`. The resulting `data.frame`, named `gs.seqs` in this example, will contain the list of full names for genes sampled in genbank for the target taxa.

```{r results = 'hide'}
gs.seqs <- gene.sampling.retrieve(organism = c("Felis", "Vulpes", "Phoca", "Manis_pentadactyla"), 
                                  speciesSampling = TRUE)
```

For the search terms used above, `phruta` was able to retrieve the names for `r nrow(gs.seqs)` gene regions. In the table below I summarize a few of those genes, with sampling frequency calculated at the level of species (see `speciesSampling = TRUE` argument above).

```{r echo=FALSE}
kable(head(gs.seqs), caption = "Gene sampling for the search including the genera Felis, Vulpes, Phoca as the ingroup taxa. Manis pentadactyla was used as outgroup.")
```

Thus, the `gene.sampling.retrieve` function provides an estimate of the number of species in genbank that matches the taxonomic criteria and have sequences for a given gene region. Note that the estimates recovered by `gene.sampling.retrieve` are only as good as the annotations that other researchers have provided for sequences deposited in genbank.

From here, we will generate a preliminary table summarizing accession numbers for the combination of taxa and gene regions that we're interested in sampling. However, note that not all these accession numbers are expected to be in the final (curated) molecular dataset. For instance, several sequences might be dropped later after taxonomic information is curated. Now, we will assemble a species-level summary of accession numbers using the `acc.table.retrieve` function. For simplicity, this tutorial will focus on sampling gene regions that are sampled in \>30% of the species (`targetGenes` `data.frame`).

```{r  results = 'hide'}
targetGenes <- gs.seqs[gs.seqs$PercentOfSampledSpecies > 30,]

acc.table <- acc.table.retrieve(
            clades  = c('Felis', 'Vulpes', 'Phoca'),
            species = 'Manis_pentadactyla' ,
            genes   = targetGenes$Gene,
            speciesLevel = TRUE
          )
```

The `acc.table` object is a `data.frame` that will be used below for downloading the relevant gene sequences. In this case, the dataset includes the following information:

```{r echo=FALSE}
row.names(acc.table) <- NULL
kable(acc.table, caption = 'Accession numbers for the relevant species and gene regions examined in this tutorial.')
```

Feel free to review this dataset, make changes, add new species, samples, etc. The integrity of this dataset is critical for the next steps so please take your time and review it carefully. For instance, let's just make some minor changes to our dataset:

```{r}
acc.table$Species <- sub("P.", "Phoca ", acc.table$Species, fixed = TRUE)
acc.table$Species <- sub("F.", "Felis ", acc.table$Species, fixed = TRUE)
acc.table$Species <- sub("V.", "Vulpes ", acc.table$Species, fixed = TRUE)
acc.table$Species <- sub("mitochondrial", "", acc.table$Species)
row.names(acc.table) <- NULL
```

Let's check how the new table looks now...

```{r echo = FALSE}
kable(acc.table, caption = 'Updated Table 2 with some minor corrections on species names. Other updates to this table include adding more sequences, using specific sequences (instead of the ones suggested by `phruta`).')
```

Now, since we're going to retrieve sequences from genbank using an existing preliminary accession numbers table, we will use the `sq.retrieve.indirect` function in `phruta`. I'm going to spend some time in here to explain the differences between the two versions of `sq.retrieve.*` in `phruta`. The one that we're using in this tutorial, `sq.retrieve.indirect`, retrieves sequences "indirectly" because it follows the initial step of generating a table summarizing accession numbers (see the `acc.table.retrieve` function above). I present the information in this vignette using `sq.retrieve.indirect` instead of `sq.retrieve.direct` because the first function is way more flexible and allows for correcting issues *prior* to download any sequence. For instance, you can add new sequences, species, populations to the resulting data.frame from `acc.table.retrieve`. Additionally, you could even manually assemble your own dataset of accession numbers to be retrieved using `sq.retrieve.indirect`. Instead, `sq.retrieve.direct` does its best to directly (i.e. without potential input from the user) retrieve sequences for a target set of taxa and set of gene regions. In short, you should be able to catch errors using `sq.retrieve.indirect` but mistakes will be harder to spot and fix if you're using `sq.retrieve.direct`. Note that the functionality of `sq.retrieve.direct` is outlined in the "Using `phruta` with defined target genes" vignette.

We still need to retrieve all the sequences from the accessions table that was generated avobe using `acc.table`. The `sq.retrieve.indirect` function will write all the resulting `fasta` files into a newly created folder `0.Sequences` located in our working directory (please check the `download.sqs = TRUE` argument).

```{r results = 'hide'}
sq.retrieve.indirect(acc.table, download.sqs = TRUE)
```

Next, we're going to make sure that we include only sequences that are reliable and from species that we are actually interested in analyzing. For this, we will be using the `sq.curate` function. We need to provide a list of taxonomic names to filter out incorrect sequences (`filterTaxonomicCriteria` argument). For simplicity, our criteria can be the genera that we're interested in analyzing. Note that the outgroup's name should also be included in the list. If the taxonomic information for a sequence retrieved from genbank does not match with any of these strings, this species will be dropped. You will have to specify whether sampling is for animals or plants (`kingdom` argument). Finally, you might have already noticed that the same gene regions can have different names. For instance, sometimes our searches retrieve both "cytochrome oxidase subunit 1" and "cytochrome c oxidase subunit I" as widely sampled genes for the target species. In that case, we can combine the sequences in these two files into a single file name `COI`. To merge gene files, you will have to provide a named list to the `mergeGeneFiles` argument of the `sq.curate` function. This named list (`tb.merged` below) will have a length that corresponds to the number of final files that should be constructed.

```{r results = 'hide'}
tb.merged <- list('COI' = c("cytochrome oxidase subunit 1", "cytochrome c oxidase subunit I"))
sq.curate(filterTaxonomicCriteria = 'Felis|Vulpes|Phoca|Manis',
          mergeGeneFiles = tb.merged,
          kingdom = 'animals', 
          folder = '0.Sequences',
          removeOutliers = FALSE)
```

Running the line of code above will create the a folder `1.CuratedSequences` containing (1) the curated sequences with original names, (2) the curated sequences with species-level names (`renamed_*` prefix), (3) a table of accession numbers (`0.AccessionTable.csv`), and (4) a summary of the taxonomic information for all the species sampled in the files (`1.Taxonomy.csv`). We'll use the `renamed_*` and `1.Taxonomy.csv` files in the next steps. Let's take a look at the sampling per gene region in the `0.AccessionTable.csv` table.

```{r echo = FALSE}
accTable <- read.csv("1.CuratedSequences/0.AccessionTable.csv")[,-1]
kable(accTable, caption = "4. Preliminary accession number table")
```

We'll now align the sequences that we just curated. For this, we just use `sq.aln` with default parameters. We need to indicate that we're interested in aligning only the `"renamed"` `fasta` files in our `1.CuratedSequences` folder.

```{r  results = 'hide'}
sq.aln(folder = '1.CuratedSequences', FilePatterns = "renamed")
```

The resulting multiple sequence alignments will be saved to the `2.Alignments` folder. In that new folder, we will have two types of files: (1) raw alignments (same file names as in `1.CuratedSequences`) and (2) alignments with ambiguous sites removed (`Masked_*` prefix). Masked alignments are only created if the `mask` argument in `sq.aln` is set to `TRUE`. In that case, one additional `.csv` file is created for each of the alignments (`0.Masked.Information_*`). Each of these datasets list the number of sites in the masked alignment that (1) are not gaps (`NonGaps` column), (2) if the sequence was removed due to the elevated number of gaps (`removedPerGaps`; controlled using the `threshold` argument in `sq.aln`), or (3) if it was removed directly in the masking step (`removedMasking`). Note that, for some gene regions, making can fail. In that case, only the original alignment file is saved to the `2.Alignments` folder.

Note that we could use these resulting alignments directly to infer our phylogenies. We cover these steps within `phruta` in another vignette: "Phylogenetics with the `phruta` R package". For now, let's wrap up and plot one of our (cool) alignments. Let's first check the raw alignments!

```{r echo=FALSE, message=FALSE, warning=FALSE}
sv <- dev.off()
```

```{r echo = FALSE}
library(ape)
files <- list.files("2.Alignments", pattern = "Raw_",full.names = TRUE)
files.main <- list.files("2.Alignments", pattern = "Raw_",full.names = FALSE)
sampleALNs <- lapply(files, read.FASTA); names(sampleALNs) <- files
par(mfrow = c(ceiling(length(sampleALNs)/2),2))
for (i in seq_along(sampleALNs)) {
  image(sampleALNs[[i]], main = files.main[i], show.labels = FALSE, xlab = "Position",
        ylab = "Species", legend = FALSE)
}
```

Now, the masked alignments...!!

```{r echo=FALSE, message=FALSE, warning=FALSE}
sv <- dev.off()
```

```{r echo = FALSE}
files <- list.files("2.Alignments", pattern = "Masked_",full.names = TRUE)
files.main <- list.files("2.Alignments", pattern = "Masked_",full.names = FALSE)
sampleALNs <- lapply(files, read.FASTA); names(sampleALNs) <- files
par(mfrow = c(ceiling(length(sampleALNs)/2),2))
for (i in seq_along(sampleALNs)) {
  image(sampleALNs[[i]], main = files.main[i], show.labels = FALSE, xlab = "Position",
        ylab = "Species", legend = FALSE)
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
sv <- dev.off()
```

```{r echo = FALSE}
end_time <- Sys.time()
```

And we're done for now!! Thanks for following this tutorial...:)

In total, this vignette took `r round(end_time-start_time)` minutes to render in my local machine. You can now try to run `phruta` using your favorite groups organisms! Don't forget to check the other tutorials and get in touch if you find any issues...Buena suerte!

```{r include = FALSE}
#Deleting the extra elements...
unlink(c("0.Sequences", "1.CuratedSequences", "2.Alignments"), recursive = TRUE)
```
